---
title: "EJ Screen Los Angeles County"
author: "Matteo Torres"
date: "December 9, 2024"
execute: 
  output: false
format: 
  html:
    theme: darkly
    toc: true
editor_options: 
  chunk_output_type: console
---

### Load Packages
```{r}
#| code-fold: true
library(here)
library(tidyverse)
```

### Import Data
```{r}
ejscreen <- read_csv(here("data", "EJSCREEN_2024_BG_StatePct_with_AS_CNMI_GU_VI.csv"))

los_angeles <- ejscreen %>%
  filter(CNTY_NAME == "Los Angeles County") %>%
  select(ID:REGION, LIFEEXPPCT, PEOPCOLORPCT, LOWINCPCT, PM25, OZONE, PTSDF) %>%
  rename_all(tolower) %>%
  drop_na()
```

### Visualiza Data (Correlation Analysis)
```{r, output = TRUE}
#| code-fold: true

# Scatterplot for Particulate Matter vs. Life Expectancy
ggplot(data = los_angeles, aes(x = pm25, y = lifeexppct)) +
  geom_point(size = 0.8) +
  labs(title = "Particulate Matter and Life Expectancy in Los Angeles County (2024)",
       x = "Particulate Matter (PM2.5)", 
       y = "Life Expectancy (%)") +
  theme_bw()
```

```{r, output = TRUE}
#| code-fold: true

# Scatterplot for Ozone vs. Life Expectancy
ggplot(los_angeles, aes(x = ozone, y = lifeexppct)) +
  geom_point(size = 0.8) +
  labs(title = "Ozone and Life Expectancy in Los Angeles County (2024)", 
       x = "Ozone", 
       y = "Life Expectancy (%)") +
  theme_bw()
```

```{r, output = TRUE}
#| code-fold: true

# Scatterplot for Proximity to Hazardous Waste Sites (PTSDF) vs. Life Expectancy
ggplot(los_angeles, aes(x = ptsdf, y = lifeexppct)) +
  geom_point(size = 0.8) +
  labs(title = "Proximity to Hazardous Waste Sites and Life Expectancy in Los Angeles County (2024)", 
       x = "Proximity to Hazardous Waste Sites (km)", 
       y = "Life Expectancy (%)") +
  theme_bw()
```

```{r}
#| code-fold: true

# Scatterplot for Percentage Low Income vs. Life Expectancy
ggplot(los_angeles, aes(x = lowincpct, y = lifeexppct)) +
  geom_point(size = 0.8) +
  labs(title = "Percentage Low Income and Life Expectancy in Los Angeles County (2024)", 
       x = "Percentage Low Income (%)", 
       y = "Life Expectancy (%)") +
  theme_bw()
```

```{r}
#| code-fold: true

# Scatterplot for Percentage People of Color vs. Life Expectancy
ggplot(los_angeles, aes(x = peopcolorpct, y = lifeexppct)) +
  geom_point(size = 0.8) +
  labs(title = "Percentage People of Color and Life Expectancy in Los Angeles County (2024)", 
       x = "Percentage People of Color (%)", 
       y = "Life Expectancy (%)") +
  theme_bw()
```

### Correlation Analysis
```{r, output = TRUE, echo = FALSE}
# Calculate correlation matrix for environmental and socio-economic variables
cor_matrix <- cor(los_angeles %>% select(pm25, ozone, ptsdf, lowincpct, lifeexppct, peopcolorpct), use = "complete.obs")
round(cor_matrix, 2)
```

|   |pm25|ozone|ptsdf|lowincpct|lifeexppct|peopcolorpct|
|---|----|-----|-----|---------|----------|------------|
|pm25|1.00|-0.10|0.59|0.17|0.12|0.46|
|ozone|-0.10|1.00|-0.52|-0.02|-0.19|0.05|
|ptsdf|0.59|-0.52|1.00|0.29|0.24|0.39|
|lowincpct|0.17|-0.02|0.29|1.00|0.41|0.52|
|lifeexppct|0.12|-0.19|0.24|0.41|1.00|0.38|
|peopcolorpct|0.46|0.05|0.39|0.52|0.38|1.00|
: Title

### Regression Models
```{r, output = TRUE}
#| code-fold: true

# Simple regression model
model <- lm(lifeexppct ~ pm25 + ozone + ptsdf, data = los_angeles)

summary(model)
```

|Coefficients|Estimate|Std. Error|t value|Pr(>|t|)|
|------------|--------|----------|-------|--------|
|(Intercept)|1.841e-01|5.296e-03|34.765|< 2e-16 ***|
|pm25|1.022e-04|3.200e-04|0.319|0.749|   
|ozone|-3.287e-04|5.364e-05|-6.128|9.43e-10 ***|
|ptsdf|7.626e-04|7.584e-05|10.056|< 2e-16 ***|
title: Simple Regression Model

Residual standard error: 0.03196 on 6289 degrees of freedom
Multiple R-squared:  0.06136,	Adjusted R-squared:  0.06091 
F-statistic:   137 on 3 and 6289 DF,  p-value: < 2.2e-16


Explanation of Model Results
Significant Predictors:
ozone and ptsdf are significant predictors (p-values < 0.001).
pm25 is not statistically significant in this model.
Adjusted R-squared:
Indicates only ~6% of the variance in life expectancy is explained by the model. This suggests other unmeasured factors might influence life expectancy.
Residual Standard Error:
~0.03196 shows the average deviation of actual values from predicted ones.

```{r, output = TRUE}
library(ggplot2)
# Reshape the data into long format for faceting
pollutants <- los_angeles %>%
  gather(key = "Pollutant", 
         value = "Value", pm25, ozone, ptsdf)  # gather predictors into one column

# Create the plot
ggplot(data = pollutants, aes(x = Value, y = lifeexppct)) +
  geom_point(size = 0.5) +
  geom_smooth(method = "lm", se = FALSE, aes(color = Pollutant)) +
  facet_wrap(~ Pollutant, scales = "free_x") +
  labs(x = "Pollutant Measure",
       y = "Life Expectancy (%)",
       title = "Life Expectancy vs. Environmental Pollutants") +
  theme_bw() +
  scale_color_manual(values = c("blue","red", "green"))
```

```{r, output = TRUE}
#| code-fold: true

# Multiple regression model
summary(lm(lifeexppct ~ pm25 + ozone + ptsdf + peopcolorpct + lowincpct, data = los_angeles))
```


|Coefficients|Estimate|Std. Error|t value|Pr(>|t|)|
|------------|--------|----------|-------|--------|
|(Intercept)|2.146e-01|4.870e-03|44.072|< 2e-16 ***|
|pm25|-1.092e-03|2.990e-04|-3.654|0.000261 ***|
|ozone|-7.951e-04|4.964e-05|-16.017|< 2e-16 ***|
|ptsdf|-1.683e-04|7.183e-05|-2.343|0.019141 *|  
|peopcolorpct|3.679e-02|1.858e-03|19.802|< 2e-16 ***|
|lowincpct|4.828e-02|2.269e-03|21.280|< 2e-16 ***|
:Multiple regression model

Residual standard error: 0.0286 on 6287 degrees of freedom
Multiple R-squared:  0.2486,	Adjusted R-squared:  0.248 
F-statistic:   416 on 5 and 6287 DF,  p-value: < 2.2e-16

```{r}
# Reshape the data into long format for faceting
coefficients <- los_angeles %>%
  gather(key = "Coefficient", 
         value = "Value", pm25, ozone, ptsdf, peopcolorpct, lowincpct)

# Create the plot
ggplot(data = coefficients, aes(x = Value, y = lifeexppct)) +
  geom_point(size = 0.5) +
  geom_smooth(method = "lm", se = FALSE, aes(color = Coefficient)) +
  facet_wrap(~ Coefficient, scales = "free_x") +
  labs( x = "Coefficient Measure",
        y = "Life Expectancy (%)",
        title = "Life Expectancy vs. Environmental Pollutants and Socioeconomic Factors") +
  theme_bw() +
  scale_color_manual(values = c("purple", "blue", "orange", "red", "green"))
```

```{r, output = TRUE}
#| code-fold: true

# Interaction model (pm25)
summary(lm(lifeexppct ~ pm25 + ozone + ptsdf + lowincpct + peopcolorpct + pm25:lowincpct + pm25:peopcolorpct, data = los_angeles))
```

|Coefficients|Estimate|Std. Error|t value Pr(>|t|)|  
|------------|--------|----------|-------|--------| 
|(Intercept)|1.762e-01|1.015e-02|17.369|< 2e-16 ***|
|pm25|2.272e-03|8.385e-04|2.710|0.006750 **|
|ozone|-8.134e-04|4.974e-05|-16.354|< 2e-16 ***|
|ptsdf|-1.960e-04|7.240e-05|-2.708|0.006796 **|
|lowincpct|7.113e-02|1.843e-02|3.860|0.000115 ***|
|peopcolorpct|8.244e-02|1.491e-02|5.528|3.38e-08 ***|
|pm25:lowincpct|-1.751e-03|1.436e-03|-1.220|0.222597|   
|pm25:peopcolorpct|-3.767e-03|1.203e-03|-3.132|0.001747 **|
:Interaction model (pm25

Residual standard error: 0.02856 on 6285 degrees of freedom
Multiple R-squared:  0.251,	Adjusted R-squared:  0.2502 
F-statistic: 300.9 on 7 and 6285 DF,  p-value: < 2.2e-16

```{r, output = TRUE}
#| code-fold: true

ggplot(los_angeles, aes(x = pm25, y = lifeexppct, color = peopcolorpct)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE, color = "red") +
  labs(x = "Particulate Matter (PM2.5)",
       y = "Life Expectancy (%)",
       title = "Interaction of PM2.5 and Percent People of Color on Life Expectancy") + 
  scale_color_gradient(low = "black", high = "orange") +
  theme_bw()
```

```{r, output = TRUE}
#| code-fold: true

ggplot(los_angeles, aes(x = pm25, y = lifeexppct, color = lowincpct)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE, color = "red") +
  labs(title = "Interaction of PM2.5 and Percent Low Income on Life Expectancy",
         x = "Particulate Matter (PM2.5)",
         y = "Life Expectancy (%)") +
  scale_color_gradient(low = "black", high = "purple") +
  theme_bw()
```

```{r, output = TRUE}
#| code-fold: true

# Interaction model (ozone)
summary(lm(lifeexppct ~ pm25 + ozone + ptsdf + lowincpct + peopcolorpct + ozone:lowincpct + ozone:peopcolorpct, data = los_angeles))
```

|Coefficients|Estimate|Std. Error|t value|Pr(>|t|)|
|------------|--------|----------|-------|--------| 
|(Intercept)|5.760e-02|1.090e-02|5.286|1.29e-07 ***|
|pm25|-3.570e-04|2.992e-04|-1.193|0.233|    
|ozone|1.046e-03|1.248e-04|8.384|< 2e-16 ***|
|ptsdf|-4.230e-04|7.293e-05|-5.800|6.96e-09 ***|
|lowincpct|3.485e-02|2.166e-02|1.609|0.108|    
|peopcolorpct|2.629e-01|1.635e-02|16.086|< 2e-16 ***|
|ozone:lowincpct|1.353e-04|2.576e-04|0.525|0.599|    
|ozone:peopcolorpct|-2.710e-03|1.937e-04|-13.991|< 2e-16 ***|
:Interaction model (ozone)

Residual standard error: 0.02804 on 6285 degrees of freedom
Multiple R-squared:  0.2781,	Adjusted R-squared:  0.2773 
F-statistic: 345.9 on 7 and 6285 DF,  p-value: < 2.2e-16


```{r, output = TRUE}
#| code-fold: true

ggplot(los_angeles, aes(x = ozone, y = lifeexppct, color = peopcolorpct)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE, color = "blue") +
  labs(title = "Interaction of Ozone and Percent People of Color on Life Expectancy",
         x = "Ozone",
         y = "Life Expectancy (%)") + 
  scale_color_gradient(low = "black", high = "orange") +
  theme_bw()
```

```{r, output = TRUE}
#| code-fold: true

ggplot(los_angeles, aes(x = ozone, y = lifeexppct, color = lowincpct)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE, color = "blue") +
  labs(title = "Interaction of Ozone and Percent Low Income on Life Expectancy",
         x = "Ozone",
         y = "Life Expectancy (%)") +
  scale_color_gradient(low = "black", high = "purple") +
  theme_bw()
```

```{r, output = TRUE}
#| code-fold: true

# Interaction model (ptsdf)
summary(lm(lifeexppct ~ pm25 + ozone + ptsdf + lowincpct + peopcolorpct + ptsdf:lowincpct + ptsdf:peopcolorpct, data = los_angeles))
```

|Coefficients|Estimate|Std.|Error|t value|Pr(>|t|)|
|------------|--------|----------|-------|--------| 
|(Intercept)|2.094e-01|5.169e-03|40.513|< 2e-16 ***|
|pm25|-1.008e-03|3.003e-04|-3.358|0.000791 ***|
|ozone|-7.771e-04|5.053e-05|-15.380|< 2e-16 ***|
|ptsdf|-1.352e-05|1.613e-04|-0.084 0.933184|   
|lowincpct|7.232e-02|6.077e-03|11.899|< 2e-16 ***|
|peopcolorpct|3.094e-02|4.551e-03|6.798|1.16e-11 ***|
|ptsdf:lowincpct|-1.178e-03|2.779e-04|-4.238|2.29e-05 ***|
|ptsdf:peopcolorpct|2.720e-04|2.281e-04|1.193|0.233109|    
:Interaction model (ptsdf)

Residual standard error: 0.02856 on 6285 degrees of freedom
Multiple R-squared:  0.2509,	Adjusted R-squared:   0.25 
F-statistic: 300.7 on 7 and 6285 DF,  p-value: < 2.2e-16

```{r, output = TRUE}
#| code-fold: true

ggplot(los_angeles, aes(x = ptsdf, y = lifeexppct, color = peopcolorpct)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE, color = "green") +
  labs(title = "Interaction of Harzardous Waste Proximity and Percent People of Color on Life Expectancy",
         x = "Hazardous Waste Proximity (km)",
         y = "Life Expectancy (%)") + 
  scale_color_gradient(low = "black", high = "orange") +
  theme_bw()
```

```{r, output = TRUE}
#| code-fold: true

ggplot(los_angeles, aes(x = ptsdf, y = lifeexppct, color = lowincpct)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE, color = "green") +
  labs(title = "Interaction of Harzardous Waste Proximity and Percent Low Income on Life Expectancy",
         x = "Harzardous Waste Proximity (km)",
         y = "Life Expectancy (%)") +
  scale_color_gradient(low = "black", high = "purple") +
  theme_bw()
```